---
title: Intro to the Tidyverse
author: Andrew Moore
date: '`r Sys.Date()`'
output:
  ioslides_presentation:
    widescreen: yes
---

## Hello!

### Currently:

  - Research Analyst with Office of Institutional Research

### Previously:

  - studied Psychology 
  - worked at University of Michigan, UM-SSW, UM ISR, UM Psychology

## General Outline

### What this talk is

- Crash Course in R & the *tidyverse*, with an emphasis on `{dplyr}`

### What this talk isn't

- Enough to make you fluent in all things R

### Part 1

- Verbs for working with data in a tidy fashion
- data: Palmer Penguins

### Part 2

- Summarizing & merging data
- data: Aggregated personality data from US States [(Rentfrow et al. (2013))](https://www.apa.org/pubs/journals/releases/psp-a0034434.pdf)

## Getting R & RStudio

<!-- more to do more to do here definitely more -->

- Installing R
  - r-project.org
  - You can download the executable for installing R from one of the mirrors, such as Microsoft's
  - https://cran.microsoft.com/

- If you want to try out R, please, please, please use RStudio!
  - RStudio is an IDE (a specialized text editor) that makes working with R (and other commonly used languages for working with data) much easier
  - https://rstudio.com/products/rstudio/download/

# Part 1 | Meet {dplyr} & the {tidyverse}

## Our First Commands: Loading Libraries & Data

- Only run once: `install.package("tidyverse")`!
- When ready to use: `library(tidyverse)`

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(palmerpenguins)

glimpse(penguins)
```

<div class="notes">
- Penguins data is loaded automatically from the package, but we'll cover importing basic data in part 2.

- Our first dplyr function: glimpse()
  - we can see the dimensions of our dataset
  - column names & data types
  - preview of what the contents of each column looks like
  
- Many functions in the tidyverse have expressive & hopefully memorable names that get across what they're for
</div>

## Using `filter()` to Subset Datasets

```{r, eval = FALSE}
# only Adelie penguins
filter(penguins, species == "Adelie")

# Adelie *OR* Gentoo
filter(penguins, species == "Adelie" | species == "Gentoo")

# this gets us the same result as above
filter(penguins, species %in% c("Adelie", "Gentoo"))

# multiple conditions can be added using commas
filter(penguins, bill_length_mm >= 45.8, year == 2007)
```

- Operators for establishing logical conditions:
  - Equal/Not equal to: `!=`, `==`
  - Greater than/less than: `>` & `<`, `>=` & `<=`
  - And/Or: `&`, `|`
  - In: `%in%`

<div class = "notes">
When our condition is based on a string, you need to wrap the entry in quotation marks
</div>

## Interlude: the Pipe Operator `%>%`

- You can read the pipe operator as "and then" in your code

- For example, a morning commute: "get in car AND THEN turn key AND THEN open garage AND THEN drive to work" using the pipe, looks like:

```
get in car %>%
    turn key %>%
    open garage %>%
    drive to work
```

```{r, eval = FALSE}
# this command...
filter(penguins, species == "Adelie")

# is the same as this!
penguins %>%
  filter(species == "Adelie")
```

# Part 2: Merging Data

## Importing Data

```{r}
library(tidyverse)

bf <- read_csv("../data/rentfrow-2013-bf-states-clean.csv")
```

## Importing Data (contd.)

```{r}
# built in lists of states & abbreviations
states <- tibble(state = state.name, state_abb = state.abb, region = as.character(state.region))

# who are we missing?
anti_join(states, bf, by = "state")
```

```{r}
# well, we won't get to learn anything about Hawaii or Alaska, bummer!
# let's add the state 'regions' onto our data
bf <- left_join(bf, states, by = "state")

# how do the regions compare as far as 'agreeableness'?
bf %>%
  group_by(region) %>%
  summarise(a_m  = weighted.mean(t_a, w = sample_size))

# hold on... why do we have missing data?
# let's find out which one isn't lining up...
filter(bf, is.na(region))

# we can fix this!
bf <- bf %>%
  mutate(
    state_abb = ifelse(state == "District of Columbia", "DC", state_abb),
    region    = ifelse(state == "District of Columbia", "Northeast", region)
  )

# now let's run our code again
bf %>%
  group_by(region) %>%
  summarise(
    a_m  = mean(t_a),
    a_sd = sd(t_a)
  )

# what if I wanted to know about the 'midwest'?
View(states)

# or:
arrange(states, region) %>% print(n = 20)

# let's use mutate() again, to create a new variable
# https://fivethirtyeight.com/features/what-states-are-in-the-midwest/
bf <- bf %>%
  mutate(
    region = ifelse(state_abb %in% c("IL", "MI", "IN", "OH", "IA"), "Midwest", region)
  )

# how many participants from each area?
bf %>%
  group_by(region) %>%
  summarise(
    total_sample = sum(sample_size),
    a_m          = mean(t_a),
    a_sd         = sd(t_a)
  )

bf %>%
  group_by(region) %>%
  summarise(
    total_sample = sum(sample_size),
    a_m          = mean(t_a),
    a_sd         = sd(t_a)
  ) %>%
  arrange(a_m)

# which 5 states are the least agreeable?
bf %>%
  arrange(t_a) %>%
  slice(1:5)

# what about *most* agreeable?
bf %>%
  arrange(desc(t_a)) %>%
  slice(1:5)
```
